---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Here we read in our data:

```{r}
data <- read.csv("C:/Users/RyanCaldwell/Documents/coursera/Reproducible_Research/Project1/activity.csv")
```

Next we want to convert the data into a data table and see the structure of table:

```{r}
library(data.table)
df <- data.table(data)
str(df)
```

We notice that the date field is not in a date format. Here we convert the date column:

```{r}
df$date <- as.Date(as.character(df$date), format="%Y-%m-%d")
```

Now check the table structure:

```{r}
str(df)
```

We also want to get an idea of how many different days of data we are dealing with. This is important for subsequent analyses in the report.

```{r}
unique(df$date)
```

Now it might be informative to know how many observations we have for each day, since that will play a part in the average.

```{r}
library(dplyr)
df %>%
  group_by(date) %>%
  summarise(n = n())
```

From this, we see there are 288 observations for each day.

## What is mean total number of steps taken per day?

Here we want to calculate the total number of steps each day:

```{r}
total_steps <- df %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))

head(total_steps)
```

Now we need to plot the results:

```{r}
library(ggplot2)
ggplot(data=total_steps, aes(total_steps)) + geom_histogram()
```

Now we need to summarize this data by mean and median:

```{r}
mean(total_steps$total_steps)
median(total_steps$total_steps)
```

## What is the average daily activity pattern?

First we need to create our dataframe of observations:

```{r}
mean_steps <- df %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))
head(mean_steps)
```

Now we need to plot the data:

```{r}
ggplot(data=mean_steps, aes(x=interval, y=avg_steps)) + 
  geom_line() +
  geom_point()
```

Now we want to know which interval, on average, has the most steps:

```{r}
mean_steps$interval[which.max(mean_steps$avg_steps)]
```

## Imputing missing values

Total number of NAs in data:

```{r}
sum(colSums(is.na(df)))
```

We will replace NAs with the average number of steps for that interval

```{r}
nas_replaced <- df
for(i in 1:nrow(nas_replaced)){
  if(is.na(nas_replaced$steps[i]) == TRUE){
    interval_type = nas_replaced$interval[i]
    x <- nas_replaced[nas_replaced$interval == interval_type,]
    interval_mean <- mean(x$interval, na.rm = TRUE)
    nas_replaced$steps[i] <- interval_mean
  }
}
```

Now let's see how the average steps after imputation compares to before imputation:

```{r}
total_steps <- nas_replaced %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))
```

```{r}
ggplot(data=total_steps, aes(total_steps)) + geom_histogram()
```

```{r}
mean(total_steps$total_steps)
median(total_steps$total_steps)
```

We see that the median only changed by around 1,000 steps, but the mean changed by a ton!


## Are there differences in activity patterns between weekdays and weekends?

We want to see if average steps is different for weekdays and weekends

```{r}
nas_replaced$weekend <- as.factor(ifelse(weekdays(nas_replaced$date) %in% c('Saturday','Sunday'), 'weekend', 'weekday'))

mean_steps <- nas_replaced %>%
  group_by(interval, weekend) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))
head(mean_steps)

ggplot(data = mean_steps, aes(x=interval, y=avg_steps)) +
  geom_line() +
  geom_point() +
  facet_wrap(~weekend, nrow=2)
```
